Learning and comparing functional connectomes across subjects

### 2013-03-26 paper.tex
Remove highlighting markers

% Define terms
``Functional connectivity'' is defined as a measure of synchronization in 
brain signals \cite{friston1994}. More generally, it is interesting as a 
window on underlying synchrony in neural processes \cite{lee2003}.
By ``functional connectome'', here we specifically denote a graph representing
functional interactions in the brain, where the term ``graph'' is taken in its
mathematical sense: a set of \emph{nodes} connected together by \emph{edges}.
Graph nodes (brain regions) correspond to spatially-contiguous and 
functionally-coherent patches of gray matter and edges describe long-range synchronizations between nodes
that are putatively subtended by large fiber pathways \cite{marrelec2006b}.  A
graph can be weighted or not, and is completely equivalent to its
\emph{adjacency matrix}, a symmetric matrix tabulating the connection weights
between each pair of nodes.  Functional-connectivity graphs are used to
represent evoked activity, as in task-response studies \cite{mcintosh2000}, as
well as ongoing activity, present in the absence of specific tasks or in the
background during task and often studied in so-called \emph{resting state}
experiments \cite{raichle2010}. Another important notion that arises from the
study of distributed modes of brain function is that of specialized functional
networks\footnote{In neuroimaging, the term network is sometimes used to denote
	a graph of brain function. To disambiguate the notion of segregated
	spatial mode \cite{fox2005} from that of connectivity graphs, we will
	purposely restrict its usage in this paper.} \cite{fox2005}. With our
definition of the functional connectome, functional networks are not directly
building blocks of the connectome but appear as a consequence of the graphical
structure \cite{varoquaux2010c,varoquaux2012}.

@@

Studying connectivity from \emph{evoked} activity with task-driven studies 
it requires disambiguating task-specific connectivity effects from
intrinsic connectivity mediated by shared neuromodulatory/task 
inputs, anatomical pathways, \emph{etc}. In this regard, it
can be beneficial to run a GLM-based first-level analysis, enforcing specificity
of the measure extracted to the task. With slow event-related designs,
task-specific functional connectivity can be captured in trial-to-trial
fluctuations in the BOLD response, estimated using a GLM analysis with one
regressor per trial \cite{grillon2012,rissman2004,mennes2010}. This approach,
known as beta-series regression, has been adapted for rapid event-related
designs, using multiple GLMs to optimize deconvolution of each trial
\cite{mumford2012}.

@@

For the problem of recovering the functional-connectivity
\emph{structure}, \emph{i.e.}\ finding which region is connected to
which, sparse inverse covariance estimators have been found to be
efficient \cite{varoquaux2010c,smith2011,ryali2012}. The intuition for relying on
inverse covariance rather than correlation stems from that fact that
standard correlation (marginal correlation) between two variables $a$ and
$b$ also capture the effects of other variables: strong correlation of
$a$ and $b$ with a third variable $c$ will induce a correlation between
$a$ and $b$. On the opposite, the inverse covariance\footnote{Covariance
and correlation matrices differ simply by the fact that a covariance
matrix captures the amplitude of a signal, via its variance, while a
correlation matrix is computed on standardized (zero mean, unit variance) signals.}
matrix (also called \emph{precision matrix} captures
partial correlations, removing the effect of other variables
\cite{marrelec2006a}. In the small sample limit, this removal is
challenging from the statistical standpoint. This is why an assumption 
of sparsity,
\emph{i.e.}\ that only few variables need to be considered at a time, is
important to estimate a good inverse covariance. Various estimation
strategies exist for sparse inverse covariance, and have an impact on the
resulting networks \cite{varoquaux2012,varoquaux2010c}. The
Graph Lasso ($\ell_1$-penalized maximum-likelihood estimator)
\cite{friedman2008} is in general a good approach for structure recovery. In group studies,
the $\ell_{21}$ estimator \cite{varoquaux2010c,honorio2012} is useful to
impose a common sparsity structure across different subjects and achieve
better recovery of this common structure. Simply put, these approaches
are necessary because estimation noise creates a background structure
(see fig.\,\ref{fig:icov_estimators}); however, unlike in a univariate situation, the
parameters are not independent, and the spurious background connections
degrade the estimation of the actual connections. The sparse estimators
make a compromise between imposing simpler models, \emph{i.e.}\ with less
connections, and providing a good fit to the data. This compromise is set via a
regularization parameter which controls the sparsity of the estimate. A
good procedure to choose this parameter is via cross-validation
\cite{varoquaux2010c}.

@@

\paragraph{Network structure extracted}
%
The correlation matrices and inverse-covariance matrices that we extract
contain a lot of information on the functional structure of the brain.
First, the correlation matrix (fig.\,\ref{fig:correlation_matrices})
shows blocks of synchronized regions that can be interpreted as
large-scale functional networks, such as the default mode network. Note
that the split in networks is not straightforward. Different ordering of the
nodes will reveal different networks. 
Indeed, because of the presence of
hubs and interleaved networks, the picture in terms of segregated networks
is not sufficient to explain full-brain connectivity
\cite{varoquaux2012}. Connectivity matrices, correlation matrices and
inverse-covariance matrices, can be represented as graphs: nodes connected
by weighted edges (inserts on fig.\,\ref{fig:correlation_matrices} and
fig.\,\ref{fig:icov_estimators}). The inverse-covariance matrix, which
captures partial correlations, appears then as extracting a
\emph{backbone} or \emph{core} of the graph.
While such structure has
been used as a way to summarize anatomical brain connectivity graphs
\cite{hagmann2008}, here it has a clear-cut meaning with regard to
the BOLD signal: it gives the conditional independence structure between
regions \cite{varoquaux2012}. In other words, regions $a$ and $b$ are
not connected if the signal that they have in common can be explained by
a third region $c$. In this light, the choice of nuisance regressors to
remove confounding common signal is less critical with partial
correlations than with correlations. Indeed, while with
correlation matrices regressing out the global mean has a drastic effect
(fig.\,\ref{fig:correlation_matrices} upper right and lower right), on
inverse covariance it only changes very slightly the resulting matrices
(fig.\,\ref{fig:icov_regressors}).

% CAMERON I really like this section, but would change it slightly to reflect
% that preprocessing has been moved to its own sectoin.
There have been debates on whether to
regress out certain signals, such as the global mean,  as it induces
negative correlations \cite{murphy2009,chang2009,fox2009}, and these may seem
surprising: one network appears as having opposite fluctuations to
another. However,
correlation between two signals only takes its meaning with the
definition of a baseline. A simple picture to explain anti-correlations
between two regions is the presence of a third region, mediating the
interactions. Using this third region as a baseline would amount to
estimate partial correlations in the whole system. Using 
inverse-covariance matrices or partial correlations to understand brain
connectivity makes the interpretation in terms of interactions between
brain regions easier and more robust to the choice of confounds.

@@

GV acknowledges funding from the NiConnect grant and the Dynamic
Diaschisis project DEQ20100318254 from \emph{Fondation pour la Recherche
M\'edicale}, as well as many insightful discussions with Andreas
Kleinschmidt on on-going activity and Bertrand Thirion on statistical
data processing.

>>>>

% Define terms
``Functional connectivity'' is defined as a measure of synchronization in 
brain signals \cite{friston1994}. More generally, it is interesting as a 
window on underlying synchrony on neural processes \cite{lee2003}.
By ``functional connectome'', here we specifically denote a graph representing
functional interactions in the brain, where the term ``graph'' is taken in its
mathematical sense: a set of \emph{nodes} connected together by \emph{edges}.
Graph nodes (brain regions) correspond to spatially-contiguous and 
functionally-coherent patches of gray matter and edges describe long-range synchronizations between nodes
that are putatively subtended by large fiber pathways \cite{marrelec2006b}.  A
graph can be weighted or not, and is completely equivalent to its
\emph{adjacency matrix}, a symmetric matrix tabulating the connection weights
between each pair of nodes.  Functional-connectivity graphs are used to
represent evoked activity, as in task-response studies \cite{mcintosh2000}, as
well as ongoing activity, present in the absence of specific tasks or in the
background during task and often studied in so-called \emph{resting state}
experiments \cite{raichle2010}. Another important notion that arises from the
study of distributed modes of brain function is that of specialized functional
networks\footnote{In neuroimaging, the term network is sometimes used to denote
	a graph of brain function. To disambiguate the notion of segregated
	spatial mode \cite{fox2005} from that of connectivity graphs, we will
	purposely restrict its usage in this paper.} \cite{fox2005}. With our
definition of the functional connectome, functional networks are not directly
building blocks of the connectome but appear as a consequence of the graphical
structure \cite{varoquaux2010c,varoquaux2012}.

@@

Studying connectivity from \emph{evoked} activity with task-driven studies 
requires disambiguating task-specific connectivity effects from
intrinsic connectivity mediated by shared neuromodulatory/task 
inputs, anatomical pathways, \emph{etc}. In this regard, it
can be beneficial to run a GLM-based first-level analysis, enforcing specificity
of the measure extracted to the task. With slow event-related designs,
task-specific functional connectivity can be captured in trial-to-trial
fluctuations in the BOLD response, estimated using a GLM analysis with one
regressor per trial \cite{grillon2012,rissman2004,mennes2010}. This approach,
known as beta-series regression, has been adapted for rapid event-related
designs, using multiple GLMs to optimize deconvolution of each trial
\cite{mumford2012}.

@@

For the problem of recovering the functional-connectivity
\emph{structure}, \emph{i.e.}\ finding which region is connected to
which, sparse inverse covariance estimators have been found to be
efficient \cite{varoquaux2010c,smith2011,ryali2012}. The intuition for relying on
inverse covariance rather than correlation stems from that fact that
standard correlation (marginal correlation) between two variables $a$ and
$b$ also capture the effects of other variables: strong correlation of
$a$ and $b$ with a third variable $c$ will induce a correlation between
$a$ and $b$. On the opposite, the inverse covariance\footnote{Covariance
and correlation matrices differ simply by the fact that a covariance
matrix captures the amplitude of a signal, via its variance, while a
correlation matrix is computed on standardized (zero mean, unit variance) signals.}
matrix (also called \emph{precision matrix}) captures
partial correlations, removing the effect of other variables
\cite{marrelec2006a}. In the small sample limit, this removal is
challenging from the statistical standpoint. This is why an assumption 
of sparsity,
\emph{i.e.}\ that only few variables need to be considered at a time, is
important to estimate a good inverse covariance. Various estimation
strategies exist for sparse inverse covariance, and have an impact on the
resulting networks \cite{varoquaux2012,varoquaux2010c}. The
Graph Lasso ($\ell_1$-penalized maximum-likelihood estimator)
\cite{friedman2008} is in general a good approach for structure recovery. In group studies,
the $\ell_{21}$ estimator \cite{varoquaux2010c,honorio2012} is useful to
impose a common sparsity structure across different subjects and achieve
better recovery of this common structure. Simply put, these approaches
are necessary because estimation noise creates a background structure
(see fig.\,\ref{fig:icov_estimators}); however, unlike in a univariate situation, the
parameters are not independent, and the spurious background connections
degrade the estimation of the actual connections. The sparse estimators
make a compromise between imposing simpler models, \emph{i.e.}\ with less
connections, and providing a good fit to the data. This compromise is set via a
regularization parameter which controls the sparsity of the estimate. A
good procedure to choose this parameter is via cross-validation
\cite{varoquaux2010c}.

@@

\paragraph{Network structure extracted}
%
The correlation matrices and inverse-covariance matrices that we extract
contain a lot of information on the functional structure of the brain.
First, the correlation matrix (fig.\,\ref{fig:correlation_matrices})
shows blocks of synchronized regions that can be interpreted as
large-scale functional networks, such as the default mode network. Note
that the split in networks is not straightforward. Different ordering of the
nodes will reveal different networks. 
Indeed, because of the presence of
hubs and interleaved networks, the picture in terms of segregated networks
is not sufficient to explain full-brain connectivity
\cite{varoquaux2012}. Connectivity matrices, correlation matrices and
inverse-covariance matrices, can be represented as graphs: nodes connected
by weighted edges (inserts on fig.\,\ref{fig:correlation_matrices} and
fig.\,\ref{fig:icov_estimators}). The inverse-covariance matrix, which
captures partial correlations, appears then as extracting a
\emph{backbone} or \emph{core} of the graph.
While such structure has
been used as a way to summarize anatomical brain connectivity graphs
\cite{hagmann2008}, here it has a clear-cut meaning with regard to
the BOLD signal: it gives the conditional independence structure between
regions \cite{varoquaux2012}. In other words, regions $a$ and $b$ are
not connected if the signal that they have in common can be explained by
a third region $c$. In this light, the choice of nuisance regressors to
remove confounding common signal is less critical with partial
correlations than with correlations. Indeed, while with
correlation matrices regressing out the global mean has a drastic effect
(fig.\,\ref{fig:correlation_matrices} upper right and lower right), on
inverse covariance it only changes the resulting matrices very slightly 
(fig.\,\ref{fig:icov_regressors}).

There have been debates on whether to
regress out certain signals, such as the global mean,  as it induces
negative correlations \cite{murphy2009,chang2009,fox2009}, and these may seem
surprising: one network appears as having opposite fluctuations to
another. However,
correlation between two signals only takes its meaning with the
definition of a baseline. A simple picture to explain anti-correlations
between two regions is the presence of a third region, mediating the
interactions. Using this third region as a baseline would amount to
estimate partial correlations in the whole system. Using 
inverse-covariance matrices or partial correlations to understand brain
connectivity makes the interpretation in terms of interactions between
brain regions easier and more robust to the choice of confounds.

@@

GV acknowledges funding from the NiConnect grant and the Dynamic
Diaschisis project DEQ20100318254 from \emph{Fondation pour la Recherche
M\'edicale}, as well as many insightful discussions with Andreas
Kleinschmidt on on-going activity and Bertrand Thirion on statistical
data processing. RCC would like to acknowledge support by a NARSAD Young
Investigator Grant from the Brain \& Behavior Research Foundation.
The authors would like to thank the anonymous reviewers
for their suggestions, which improved the manuscript.
